Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-28 paper.qmd
r2u and portable binaries

While comprehensive dependency tracking could theoretically be implemented in
any package manager, what distinguishes [Nix]{.pkg} (alongside its cousin
[Guix]{.pkg}) is its *functional package management* paradigm. Traditional
package managers require imperative, step-by-step installation instructions: *do
this, then this, then that*. [Nix]{.pkg} inverts this model; users declare the
desired end state (*I require an environment with these packages*), and
[Nix]{.pkg} determines how to achieve it.

This is possible because [Nix]{.pkg} packages are defined as *derivations*:
declarative blueprints specifying all inputs (source code, build commands,
dependencies) needed to build a package. When installing software, [Nix]{.pkg}
evaluates expressions written in the [Nix]{.proglang} language to produce these
derivations. The declarative nature ensures every build process is fully
specified and independent of the execution environment.

For example, the [rJava]{.pkg} package definition includes the correct
`JAVA_HOME` configuration pointing to the [jdk]{.pkg} dependency. Outside
[Nix]{.pkg}, users must manually set this variable if automatic detection fails.
With [Nix]{.pkg}, the package maintainers handle this complexity once (by
writing the correct [Nix]{.pkg} expressions that perform the installation
purely), and all users benefit.

[Nix]{.pkg} conceptualizes software builds as *pure functions* from functional
programming. In mathematics, a pure function always produces the same output for
a given input: if $f(x) = x^2$, then $f(2) = 4$ invariably holds. Analogously,
[Nix]{.pkg} ensures that identical inputs (source code, dependencies, build
instructions) produce identical outputs (built packages), regardless of when or
where the build occurs.

>>>>

Comprehensive dependency tracking, however, is not unique to [Nix]{.pkg}. The
[r2u]{.pkg} project by @r2u achieves something similar for Ubuntu: once the r2u
repository is configured, every [R]{.pkg} package is installed as a binary with
all its dependencies automatically resolved. In principle, any package manager
could implement such tracking.

What distinguishes [Nix]{.pkg} (and its sibling [Guix]{.pkg}, the only other
functional package manager) is its *functional package* management model. This
model makes full dependency specification intrinsic to the system, mandatory by
design, and replaces the imperative, step-by-step nature of traditional package
managers (*install this, then that*) with a declarative one: users describe the
desired end state (*an environment with these packages*), and [Nix]{.pkg}
determines how to realize it.

A further advantage is that packages built by [Nix]{.pkg} act as universal
binaries across any Linux system where [Nix]{.pkg} is installed. This “build
once, run anywhere” property has long been solved by [Nix]{.pkg}, and is now
being revisited by Posit through its initiative to provide portable CRAN
binaries [@posit_portable_r_binaries].

This universality stems from [Nix]{.pkg}’s use of derivations, declarative
blueprints that specify all inputs (source code, build commands, dependencies)
required to build a package. When software is installed, [Nix]{.pkg} evaluates
expressions written in the [Nix]{.proglang} language to produce these
derivations. Because every input is explicitly declared, builds are fully
reproducible and independent of the surrounding environment.

For example, the [rJava]{.pkg} package definition includes the correct
`JAVA_HOME` configuration pointing to its [jdk]{.pkg} dependency. Outside
[Nix]{.pkg}, users must manually set this variable if automatic detection fails.
Within [Nix]{.pkg}, maintainers encode such details once, ensuring every user
benefits from a consistent, reliable installation.

At its core, [Nix]{.pkg} treats software builds as pure functions, a concept
borrowed from functional programming. A pure function always produces the same
output for a given input—for example, if $f(x) = x^2$, then $f(2) = 4$
invariably holds. Likewise, [Nix]{.pkg} guarantees that identical inputs (source
code, dependencies, build instructions) yield identical outputs (built
packages), regardless of when or where the build occurs.
